---
title: 'Warmup Exercise: refactoring prepIJDF function'
author: "Mae Hutchison"
date: "2026-05-30"
output:
  html_document:
    toc: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bench)
library(ggplot2)
library(ggbeeswarm)

prev_prepIJDf <- function(df, mij=NULL, dij=NULL, logTrans=FALSE, zTrans=FALSE){
  ##############################################################################
  ## Indices:
  ##           '.i'    for origin 
  ##           '.j'    for destination
  ##           '.ij'   for origin/destination ratio
  ##           '.ijF'  factor variable
  ##           '.ijL'  log-transformed variable
  ##           '.ijLZ' log- and z-transformed variable
  ##           '.ijZ'  z-transformed variable
  ## Case treatments:
  ##            1 Factor
  ##            2 NoFactor-possitive
  ##            3 NoFactor-possitive-log
  ##            4 NoFactor-possitive-log-scale
  ##            5 NoFactor-possitive-scale
  ##            6 NoFactor
  ##            7 NoFactor-scale
  ##############################################################################
  if(!is.data.frame(df)) stop("Function prepIJDf: Input not a dataframe")
  
  ## Setup origin and destination ids and unique record id with the format: 
  ## 'ij' for n < 10, iijj' for n < 100 and 'iiijjj' for n < 1000 etc.
  recId <- expand.grid(1:nrow(df),1:nrow(df))
  odDf <- data.frame(ID.i=recId$Var1, ID.j=recId$Var2,
                     ID.ij=recId$Var1*10^trunc(log(nrow(df),base=10)+1)+recId$Var2)
  
  ## Add vectorized migration flows
  if (is.null(mij)) { odDf <- data.frame(odDf, mij=NA) } else {
    odDf <- data.frame(odDf, mij=matrix(as.matrix(mij, nrow=nrow(mij)^2)) ) }
  
  ## Add vectorized distances
  if (is.null(dij)) { odDf <- data.frame(odDf, dij=NA) } else {
    odDf <- data.frame(odDf, dij=matrix(as.matrix(dij, nrow=nrow(dij)^2)) ) }
  
  ## Cycle over all variables
  dfNames <- names(df)  
  for (i in 1:length(dfNames)) {
    odVec <- expand.grid(df[,i],df[,i])
    
    if (is.factor(df[,i])) {                              # process factors
      odVar <- data.frame(odVec[,1],odVec[,2])
      names(odVar) <- c(paste0(dfNames[i],".iF"), paste0(dfNames[i],".jF"))
      odVar[,1] <- factor(odVec[,1], labels=levels(df[,i]))
      odVar[,2] <- factor(odVec[,2], labels=levels(df[,i]))
    } else { 
      if(!(any(df[,i] <= 0))){                      # process all metric positive variables                                
        if (!logTrans & !zTrans){
          odVar <- data.frame(odVec[,1],odVec[,2],log( odVec[,1]/odVec[,2]))
          names(odVar) <- c(paste0(dfNames[i],".i"),paste0(dfNames[i],".j"),
                            paste0(dfNames[i],".ijL"))
        } 
        if (logTrans & !zTrans){
          odVar <- data.frame(log(odVec[,1]),log(odVec[,2]),log(odVec[,1]/odVec[,2]))
          names(odVar) <- c(paste0(dfNames[i],".iL"),paste0(dfNames[i],".jL"),
                            paste0(dfNames[i],".ijL"))        
        }
        if (logTrans & zTrans){
          odVar <- data.frame(log(odVec[,1]),log(odVec[,2]),log( odVec[,1]/odVec[,2]))
          odVar <- as.data.frame(scale(odVar))
          names(odVar) <- c(paste0(dfNames[i],".iLZ"),paste0(dfNames[i],".jLZ"),
                            paste0(dfNames[i],".ijLZ"))           
        }
        if (!logTrans & zTrans){
          odVar <- data.frame(odVec[,1],odVec[,2],log( odVec[,1]/odVec[,2]))
          odVar <- as.data.frame(scale(odVar))
          names(odVar) <- c(paste0(dfNames[i],".iZ"),paste0(dfNames[i],".jZ"),
                            paste0(dfNames[i],".ijLZ")) 
        }           
      } else {
        if (!zTrans){
          odVar <- data.frame(odVec[,1],odVec[,2])
          names(odVar) <- c(paste0(dfNames[i],".i"),paste0(dfNames[i],".j"))             
        }
        if (zTrans){
          odVar <- data.frame(odVec[,1],odVec[,2])
          odVar <- as.data.frame(scale(odVar))
          names(odVar) <- c(paste0(dfNames[i],".iZ"),paste0(dfNames[i],".jZ"))}
      }##end::variable support
    } ##end::isFactor
    odDf <- data.frame(odDf, odVar)
  } #end::for
  return(odDf)
} #end::prev_prepIJDf
```

# Task
Refactor the `prepIJDf` function from the Italy Migration example from GISC 7310 (week 12).

_Priorities:_  
- keep as an `s3` method  
- minimize package dependencies (ideally none)  
- code should be easy to follow for students seeing it for the first time  
- code should be structured in a way that is easy to maintain/make future updates  


# Refactored prepIJDF function  

The refactored function takes the same inputs and returns the same outputs as the 
original function. The only added functionality is some additional stop conditions
that provide a descriptive error message to the student if an invalid argument 
is passed (also prevents unexpected behavior from unanticipated inputs).
No dependencies were added.

_Optimization_: defined a helper function `expand_column_to_od` that can be applied
to the input dataframe using `lapply`, avoiding the `for` loop of the original function.

_Organization_: condensed the transformation logic to `if`/`else` statement at the 
beginning to improve readability. Also moved `ID.ij` assignment  to helper 
function `make_pair_id` to provide additional details in comments without cluttering
the main function; also allows students to play around with the pair id function by
itself to better understand it.

_Maintainability_: changed a few `if`/`else` blocks to assignment statements
(e.g., now `$mij` is assigned only once overall, instead of once in each condition block).
Also changed any un-ordered, multi-line lists to delimiter first (leading comma) 
for cleaner version control diffs.

_Style_: Changed all camel case varnames to snake case (except for input options).

_Documentation_: expanded the function description comment for clarity.  

```{r new_prepIJDf}
new_prepIJDf <- function(df, mij=NULL, dij=NULL, logTrans=FALSE, zTrans=FALSE){
  ##############################################################################
  ## Build a long-format origin-destination (OD) dataframe from a wide-format
  ## input dataframe, expanding every variable into origin (.i), destination (.j)
  ## and log-ratio (.ijL) columns for all n^2 OD pairs.
  ##
  ## Arguments:
  ##   df        - dataframe with n rows, one per location. Each column is
  ##               expanded into OD pair format.
  ##   mij       - optional n x n list of migration flows. If NULL,
  ##               the mij column is set to NA.
  ##   dij       - optional n x n numeric matrix of distances or costs between
  ##               locations. If NULL, the dij column is omitted.
  ##   logTrans  - logical boolean flag indicating if numeric variables are to 
  ##               be log-transformed.
  ##   zTrans    - logical boolean flag indicating if numeric variables are to
  ##               be standardized (scale of -1 to 1)
  ## Indices:
  ##           '.i'    for origin 
  ##           '.j'    for destination
  ##           '.ij'   for origin/destination ratio
  ##           '.ijF'  factor variable
  ##           '.ijL'  log-transformed variable
  ##           '.ijLZ' log- and z-transformed variable
  ##           '.ijZ'  z-transformed variable
  ##############################################################################
  stopifnot( 
     "df must be a dataframe" = is.data.frame(df)
    ,"logTrans must be a logical boolean" = is.logical(logTrans)
    ,"zTrans must be a logical boolean" = is.logical(zTrans)
    ,"mij must be a matrix or data frame of size n x n" =
        is.null(mij) ||
        ((is.matrix(mij) || is.data.frame(mij)) && all(dim(mij) == nrow(df)))
    ,"dij must be a numeric array of size n x n"= 
        is.null(dij) ||
        (is.double(dij) && all.equal(dim(dij),rep(nrow(df),2)))
  )
  
  transform <- if(logTrans&&zTrans){
    "log.standard"
  }else if(logTrans){
    "log"
  }else if(zTrans){
    "standard"
  }else{
    "none"
  }
  
  ## Setup origin and destination ids and unique record id with the format: 
  ## 'ij' for n < 10, iijj' for n < 100 and 'iiijjj' for n < 1000 etc.
  pairs <- expand.grid(i = 1:nrow(df), j = 1:nrow(df))
  od_df <- data.frame(
    ID.i  = pairs$i,
    ID.j  = pairs$j,
    ID.ij = make_pair_id(pairs$i, pairs$j)
  )
  
  ## Add vectorized migration flows
  od_df$mij <- if (is.null(mij)) {
    NA
  } else {
    as.vector(as.matrix(mij))
  }
  
  ## Add vectorized distances
  od_df$dij <- if(is.null(dij)){
    NA
  } else {
    as.vector(as.matrix(dij, nrow=nrow(dij)^2))
  }
  
  ## Cycle over all variables
  od_vars <- lapply(names(df), expand_column_to_od, df = df, transform=transform)
  
  ## Combine all variables into one data frame
  od_df    <- do.call(data.frame, c(list(od_df), od_vars))
  
  return(od_df)
} #end::prepIJDf
```

## Helper functions
```{r helper_functions}
make_pair_id <- function(i, j, n=max(c(i, j))) {
  ##############################################################################
  ## Helper function to create pair IDs of every combination of vectors i and j
  ## i, j are integer vectors
  ## n: the upper bound of i & j, used to set digit-width so IDs don't collide.
  ##    Default max(i,j) but should be set explicitly when i/j may be a subset
  ##    of their full range.
  ## ex: 'ij' for n < 10, iijj' for n < 100 and 'iiijjj' for n < 1000 etc.
  ##############################################################################
  stopifnot(  "i must be an integer vector"=is.integer(i)
            , "j must be an integer vector"=is.integer(j)
            , "n must be a single positive integer"
               =(is.integer(n) && (n>0) && length(n) == 1)
            )
  digits <- trunc(log10(n) + 1)
  return(i * 10^digits + j)
}# end::make_pair_id


expand_column_to_od <- function(col_name, df, transform) {
  ##############################################################################
  ## Helper function to process one column of a dataframe to ij format
  ## col_name is the column name corresponding to column in df to process
  ## df is the data frame with the source data to process
  ## transform is a string 'log', 'standard', 'log.standard', or 'none'
  ##############################################################################
  stopifnot("transform must be str 'log', 'standard', 'log.standard', or 'none'" 
            = transform %in% c("log", "standard", "log.standard", "none")
            , "df must be a data frame" = is.data.frame(df)
            , "col_name muse be a column name in df" = col_name %in% names(df)
            )
  
  col    <- df[[col_name]] # column of interest
  od_vec <- expand.grid(col, col) # origin-destination vector
  
  if (is.factor(col)) { # for factor columns, only make ij pairs
    col_levels <- levels(df[[col_name]])
    od_var <- data.frame(
      i = factor(od_vec[, 1], labels = col_levels),
      j = factor(od_vec[, 2], labels = col_levels)
    )
    names(od_var) <- c(paste0(col_name, ".iF"), paste0(col_name, ".jF"))
  } else { # for numeric columns, make ij pairs and log ratio, 
           # and apply any transformations
    i_col  <- od_vec[, 1]
    j_col  <- od_vec[, 2]
    is_log      <- transform %in% c("log", "log.standard")
    is_standard <- transform %in% c("standard", "log.standard")
    all_positive <- !any(df[[col_name]] <= 0)
    
    od_var <- if (all_positive) {
      data.frame(
        i   = if (is_log) log(i_col) else i_col,
        j   = if (is_log) log(j_col) else j_col,
        ratio = log(i_col / j_col)
      )
    } else {
      # non-positive numeric variables will only generate ij pairs, 
      # ignores any log transformation setting passed to `transform`
      data.frame(i = i_col, j = j_col)
    }
    
    if (is_standard) {
      od_var <- as.data.frame(scale(od_var))
    }
    
    suffix <- switch(transform,
                     "log"          = c(".iL",  ".jL",  ".ijL"),
                     "standard"     = c(".iZ",  ".jZ",  ".ijLZ"),
                     "log.standard" = c(".iLZ", ".jLZ", ".ijLZ"),
                     "none"         = c(".i",   ".j",   ".ijL")
    )
    
    names(od_var) <- paste0(col_name, suffix[seq_len(ncol(od_var))])
  } # end::numeric columns
  
  return(od_var)
} # end::expand_column_to_od
```

## Performance
```{r make_synthetic_inputs, echo=FALSE}
make_synthetic_inputs <- function(n = 95) {
  
  df <- data.frame(
    pop      = as.double(sample(1000:1000000, n, replace = TRUE)),
    gdp      = runif(n, min = 1e6, max = 1e10),
    density  = runif(n, min = 10, max = 5000),
    unemployment = runif(n, min = 2, max = 25),
    region   = factor(sample(c("north", "south", "central"), n, replace = TRUE)),
    coastal  = factor(sample(c("yes", "no"), n, replace = TRUE))
  )
  
  # mij: migration flows as a data frame 
  mij <- outer(df$pop, df$pop, function(i, j) {
    as.integer(i * j / 1e10 * runif(1))
  })
  mij <- as.data.frame(mij)
  
  # dij: distances as a numeric matrix
  coords <- data.frame(
    x = runif(n, 0, 1000),
    y = runif(n, 0, 1000)
  )
  dij <- as.matrix(dist(coords))
  
  list(df = df, mij = mij, dij = dij)
}
```

The vectorized function performs slightly better and had fewer garbage collections (gc). 

```{r performance_check, cache=TRUE}
set.seed(37)
inputs <- make_synthetic_inputs(n = 100)
df  <- inputs$df
mij <- inputs$mij
dij <- inputs$dij

results <- bench::mark(
  previous = prev_prepIJDf(df, mij = mij, dij = dij, logTrans = TRUE, zTrans = TRUE),
  new = new_prepIJDf(df, mij = mij, dij = dij, logTrans = TRUE, zTrans = TRUE),
  iterations = 100,
  check = TRUE
)

print(results)
plot(results) 
```

# Alternative: build_od_pairs
I also made a modified version the `prepIJDf` function called `build_od_pairs`.
This function removes the `ID.ij` index (forces student to use both `ID.i` and `ID.j` 
columns to reference unique rows), and sets rows where `i==j` to `NA` (throws an 
error if student forgets to account for diagonal elements, except for in `glm`
functions where now the correct behavior (ignoring the diagonals) happens by 
default due to `na.action` being set to omit `NA`).

It also simplifies the transformation option input by passing a descriptive 
string instead of combo of two booleans.

```{r}
build_od_pairs <- function(df, mij = NULL, dij = NULL, transform = "none") {
  ##############################################################################
  ## Build a long-format origin-destination (OD) dataframe from a wide-format
  ## input dataframe, expanding every variable into origin (.i), destination (.j)
  ## and ratio (.ij) columns for all n^2 OD pairs.
  ##
  ## Arguments:
  ##   df        - dataframe with n rows, one per location. Each column is
  ##               expanded into OD pair format.
  ##   mij       - optional n x n numeric matrix of migration flows. If NULL,
  ##               the mij column is set to NA.
  ##   dij       - optional n x n numeric matrix of distances or costs between
  ##               locations. If NULL, the dij column is omitted.
  ##   transform - character string controlling variable transformation.
  ##               One of:
  ##                 "none"         : no transformation (default)
  ##                 "log"          : log-transform i, j, and ij columns
  ##                 "standard"     : z-standardize i, j, and ij columns
  ##                 "log.standard" : log-transform then z-standardize
  ##
  ## Returns:
  ##   A dataframe with n^2 rows, one per OD pair, containing:
  ##     ID.i, ID.j  - integer indices of origin and destination
  ##     mij         - migration flow (if provided)
  ##     dij         - distance/cost (if provided)
  ##     <var>.i     - origin value for each variable in df
  ##     <var>.j     - destination value for each variable in df
  ##     <var>.ijL    - log ratio of origin to destination (positive vars only)
  ##    * all od pairs where i=j have a value set to NA to prevent accidental use*
  ##
  ## Column suffixes:
  ##   .i    origin value
  ##   .j    destination value
  ##   .ijL  log ratio (origin / destination)
  ##   .iF   origin factor
  ##   .jF   destination factor
  ##   .iL   log-transformed origin
  ##   .jL   log-transformed destination
  ##   .iZ   z-standardized origin
  ##   .jZ   z-standardized destination
  ##   .iLZ  log- and z-transformed origin
  ##   .jLZ  log- and z-transformed destination
  ##   .ijLZ log- and z-transformed destination
  ##############################################################################
  stopifnot("df must be a dataframe" = is.data.frame(df)
            ,"mij must be a numeric matrix of size n x n" 
             = is.null(mij) ||
               (is.matrix(mij) && all.equal(dim(mij),rep(nrow(df),2)))
            ,"dij must be a numeric matrix of size n x n"
             = is.null(dij) ||
               (is.matrix(dij) && all.equal(dim(dij),rep(nrow(df),2)))
            ,"transform must be str 'log', 'standard', 'log.standard', or 'none'" 
             = transform %in% c("log", "standard", "log.standard", "none")
  )
  
  
  ## Setup origin and destination ids 
  pairs <- expand.grid(i = 1:nrow(df), j = 1:nrow(df))
  od_df <- data.frame(ID.i=pairs$i, ID.j=pairs$j)
  
  ## Add vectorized migration flows
  od_df$mij <- if (is.null(mij)) {NA} else {as.vector(mij)}
  
  ## Add vectorized distances
  od_df$dij <- if(is.null(dij)){NA} else {as.vector(dij)}
  
  ## Apply expansion over all variables
  od_vars <- lapply(names(df), expand_column_to_od, df = df, transform=transform)
  
  ## Combine all expanded variables into one data frame
  od_pairs    <- do.call(data.frame, c(list(od_df), od_vars))
  
  ## Set variables for diagonal (i == j) pairs to NA
  self_pairs <- od_pairs$ID.i == od_pairs$ID.j
  var_cols    <- !names(od_pairs) %in% c("ID.i", "ID.j", "mij", "dij")
  od_pairs[self_pairs, var_cols] <- NA
  
  return(od_pairs)
} #end::build_od_pairs
```

## Performance
This function performs a little slower than the refactored `prepIJDf` function
(possibly due to extra string comparison and diagonal var assignment); however,
it is still vectorized and is realtively fast.

```{r performance_2, cache=TRUE}
results <- bench::mark(
  "original prepIJDF" = prev_prepIJDf(df, mij = mij, dij = dij
                                      , logTrans = TRUE, zTrans = TRUE),
  "refactored prepIJDf" = new_prepIJDf(df, mij = mij, dij = dij
                                       , logTrans = TRUE, zTrans = TRUE),
  "build_od_pairs" = build_od_pairs(df, mij = as.matrix(mij), dij = dij, transform = "log.standard"),
  iterations = 100,
  check = FALSE
)

print(results)
plot(results) 
```


# Testing

Test cases are in `00_warmup/tests/testthat`.
```{r testing, results='markup'}
setwd("~/GitHub_m-hutch/independent-study-su26")
source("00_warmup/tests/testthat.R")
```

# Notes / Questions

Both functions follow the behavior of the original `prepIJDf` of centering and 
scaling the entire data frame, including the log-ratio columns, when `zTrans`
is true (or `transform` is "standard" or "log.standard"). I'm not sure if this 
would always be the desired behavior, since scaling the log-ratio column makes
it much more difficult to interpret.
