---
title: "TexMix"
subtitle: "An R Package for Teaching Demographic, Migration, and Spatial Data Analysis"
author: "Mae Hutchison"
date: last-modified
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    toc: false
    code-fold: false
execute:
  echo: true
  output: true
---

## Agenda

I. Introduction  
II. Literature Review  
III. Study Area and Data Sources  
IV. Analysis / Methodology  
V. Results and Discussion  
VI. Conclusions and Future Study  
VII. References / Bibliography  

# I. Introduction {background-color="#1b3a5c"}

## What is TexMix?

- An `R` package built to support **teaching spatial demographic analysis** at UT Dallas
- Bundles curated Census/ACS data, spatial data frames, and purpose-built functions
- Designed for teaching and student's own applied research 
- Goal: provide a comprehensive package for teaching courses that use `R`, with easily interpretable code that students can use as a reference when writing their own scripts

## Why TexMix?

- Off-the-shelf Census workflows require repetitive boilerplate: pulling ACS variables, joining geometries, managing GEOIDs
- Migration and flow data are especially painful to reshape for mapping and vector-based analysis
- TexMix pre-packages the data and functions our students and collaborators use most often
- This release focuses on **new data coverage**, **new analytical functions**, and **performance / packaging improvements**

## Release Highlights

::: {.incremental}
- New 2024 ACS data for NCTCOG tracts and blockgroups, plus new factor variables and new 2024 ACS data for all Texas counties
- Expanded highway and water spatial layers and added a new Texas neighbors `sdf` for basemaps
- New Italy migration dataset and basemap files
- Three new functions: `flow2vec`, `ageStructure`, `viewPaletteSpread`; updated `mapColorRamp`
- Vectorized performance improvements, universal GEOID keys, and a new testing framework
:::

# II. Literature Review {background-color="#1b3a5c"}

## Foundations in Spatial Demography

- Long tradition of using Census geography (tracts, block groups, counties) as the building blocks of demographic analysis
- R's spatial ecosystem has matured into the de facto standard for reproducible Census workflows
- Migration and commuting flow analysis has its own lineage: origin-destination matrices, desire-line mapping, and flow vectorization techniques used in transportation and migration research

## Positioning TexMix

- TexMix builds on, rather than replaces, the `sp` / `sf` ecosystem
- Distinguishing contributions:
  - Pre-joined, analysis-ready ACS extracts for a defined study region (NCTCOG) and for Texas statewide
  - Purpose-built helpers for **flow data** (`flow2vec`) and **age-structure diagrams** (`ageStructure`) that are otherwise assembled ad hoc
  - Cartographic helper functions (`mapColorRamp`, `viewPaletteSpread`) aimed at consistent, accessible thematic mapping
- Related work: regional council data portals (NCTCOG), Census Bureau flow reports, and prior TexMix releases that established the base package structure

## Gaps TexMix Addresses

- Regional (NCTCOG) data is published separately from US Census and statewide Texas data -- TexMix unifies both under a common GEOID schema
- Flow/OD data preparation for vector-based mapping is typically hand-coded per project -- `flow2vec` generalizes this step
- Color palette selection for choropleth maps is rarely tested for perceptual spread -- `viewPaletteSpread` gives users a diagnostic tool
- Department-specific teaching tool -- fully customized to fit UTD's spatial analysis courses

# III. Study Area and Data Sources {background-color="#1b3a5c"}

## Study Area

- **Primary region:** Dallas County, with buffer data for all North Central Texas Council of Governments (NCTCOG) counties
  - Tract- and blockgroup-level detail
- **Secondary region:** All 254 Texas counties (statewide generalization)
- **For migration vignette** Italy migration data, included for cross-national migration comparison exercises
```{r}
library(TexMix)
```


## DFW Region -- Census Tracts
```{r sa1}
data("DFW_tractShp")
plot(DFW_tractShp)
```

## DFW Region --  Block Groups
```{r}
data("DFW_bgShp")
plot(DFW_bgShp)
```

## Texas Counties
```{r}
data("TX_countyShp")
plot(TX_countyShp)
```

## Italy provinces
```{r}
data("ItalyShp")
plot(ItalyShp)
```


## New Data -- NCTCOG Tracts and Blockgroups

- 2024 ACS data for census tracts and blockgroups within NCTCOG counties
- New factor variables added to support classification and grouping:
  - **School district**
  - **Municipality**
  - **Congestion level**
- These factors enable direct filtering, faceting, and choropleth mapping without additional joins

## Census data
```{r}
colnames(DFW_tractShp@data)
```
## New Factor Data Example
```{r}
mapColorQual(DFW_tractShp@data$CongLevel, DFW_tractShp)
```


## New Data -- Statewide and Spatial Layers

- 2024 ACS data extended to **all Texas counties**, enabling state-level context alongside the NCTCOG focus
- **Expanded highway and water layers** for richer basemap cartography and spatial reference
- **Texas neighbors sdf**: a spatial data frame of bordering states, for cleaner basemaps on any Texas-extent map
- **Italy migration data and basemap files**: adds an international migration dataset for comparative coursework and research

## Example -- Texas neighbors
```{r}
data("TX_neighShp")
extent <- bbox(TX_countyShp)
plot(TX_neighShp, xlim=extent[1,], ylim=extent[2,])
mapColorRamp(TX_countyShp, "POPDEN", add.to.map = T, legend.pos="bottomright")
```


## Data Design Principle: Universal GEOID

- Every new dataset in this release includes a **GEOID** field
- Rationale:
  - Consistent join key across tracts, blockgroups, and counties
  - Simplifies adding future variables without re-engineering joins
  - Enables long-term maintainability as ACS vintages update annually

# IV. Analysis / Methodology {background-color="#1b3a5c"}

## New Function: `flow2vec`

- Converts origin-destination / migration flow tables into vector-ready formats for mapping and analysis
- Formerly implemented as `prepIJDf`in an in-class script; renamed and substantially rebuilt
- **Performance improvement:** parallel calculations across columns are now vectorized, rather than looped, reducing processing time on large flow tables

## Example: flow2vec
```{r}
data("ItalyShp")
data("Italy_migration")

df <- ItalyShp@data[,c("TOTPOP94","TOTFERTRAT")]
mij <- as.matrix(Italy_migration[,2:96])
vec.migration <- flow2vec(df, mij=mij)
head(vec.migration)
```


## New Function: `ageStructure`

- Generates standardized age-structure diagram (population pyramid) from ACS age variables
- Reduces repetitive recoding of age bins across projects
- Designed to work directly with the new NCTCOG and statewide ACS extracts

## Exampe: ageStructure

```{r}
ageStructure(TX_countyShp, name="Texas")
```


## New Function: `viewPaletteSpread`

- Diagnostic tool for evaluating color palettes before mapping
- Helps users visually check perceptual spacing/contrast of a palette across classes
- Complements the updated `mapColorRamp` function by letting users see how observations are distributed within the palette

## Example: viewPaletteSpread
```{r include=FALSE}
ramp <- mapColorRamp(DFW_tractShp, 'POPDEN', breaks=5, output.breaks=TRUE)
```
```{r}
viewPaletteSpread(ramp)
```


## Updated Function: `mapColorRamp`

- Refined to provide more palette options and user-control over the plot
```{r}
mapColorRamp(ItalyShp, 'TOTPOP94', breaks=5, palette = "Blues")
```


## Methodological / Technical Updates

- **Vectorization:** `flow2vec` parallelizes calculations across columns instead of row-wise iteration, improving performance on large datasets
- **GEOID standardization:** all new datasets carry GEOID for easy joins and future variable additions
- **Testing framework:** new functions (`flow2vec`, `ageStructure`, `viewPaletteSpread`) are now covered by automated tests, supporting long-term reliability and CRAN submission readiness

# V. Results and Discussion {background-color="#1b3a5c"}

## What This Release Enables

- Analysts can now move between **regional (NCTCOG)** and **statewide (Texas)** scales using the same GEOID-based join logic
- Migration/commuting flow work is faster and less error-prone with `flow2vec`'s vectorized approach
- Age-structure analysis and thematic color mapping are now first-class, tested functions rather than ad hoc scripts
- The new Texas neighbors sdf and expanded highway/water layers improve basemap quality for any Texas-focused map


## Known Considerations

- New factor variables (school district, municipality, congestion level) can shift year to year
- Test coverage should be expanded to include existing functions, not just the three new functions
- old data needs to be phased out of the package

# VI. Conclusions and Future Study {background-color="#1b3a5c"}

## Conclusions

- This release meaningfully expands TexMix's data footprint (NCTCOG + statewide Texas + Italy) and analytical toolkit (`flow2vec`, `ageStructure`, `viewPaletteSpread`)
- Technical investments -- vectorization, universal GEOID, and a testing framework -- position the package for more sustainable, reliable growth
- TexMix is closer to being a robust, general-purpose teaching and research tool for Texas spatial demography

## Next Steps

::: {.incremental}
- **Submit final package to CRAN**
- **Set up an official GitHub repository**
- **Build compiled tarballs for all supported systems**
- **Gather feedback from students** using the package in coursework
:::

## Future Study

- Expand the testing framework to cover all exported functions, including updated ones
- Annual ACS refresh workflow, leveraging the new GEOID standardization
- Long term: will need to migrate to `sf` package once `sp` is decomissioned

# VII. References / Bibliography {background-color="#1b3a5c"}

## References

- U.S. Census Bureau. *American Community Survey (ACS) 2024, 5-Year Estimates.*
- North Central Texas Council of Governments (NCTCOG). *Regional Data and GIS Resources.*
- Pebesma, E. (2018). *Simple Features for R: Standardized Support for Spatial Vector Data.* The R Journal.
- Walker, K. (2023). *Analyzing US Census Data: Methods, Maps, and Models in R.*
- TexMix package documentation and prior release notes (internal).

## Thank You

- Questions and feedback welcome
- Package repository, documentation, and CRAN listing forthcoming per the next-steps roadmap
