---
title: "Dallas Market Area Analysis with Contiguity Constraint"
author: "Michael Tiefelsdorf"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: yes
#output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dallas Market Areas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "R>", cache=FALSE, fig.width=7.2)
rm(list=ls())                                        # start fresh
library(TexMix)                    # Get data and support function plotBoxesByFactor.r
library(ClustGeo)                  # Hierarchical Cluster Analysis with Spatial Constraint
library(sp)
library(VIM)
library(spdep)
library(RColorBrewer)
```
# Introduction

The objective of this __prototype cluster analysis__ it to perform unsupervised 
learning to identify distinct and homogeneous market areas in Dallas County with 
respect to their 

* socio-economic, 
* demographic and 
* infrastructure 

characteristics of the underlying census tracts.

The package "ClustGeo" is the workhorse of this analysis. It performs hierarchical
cluster analysis using the Ward parametrization of the Lance & Williams algorithm
under an imposed spatial contiguity constraint.

Potential improvements could be:

* Include additional features of the census tracts to the analysis. However, 
  maintain a balance between the socio-economic, demographic and infrastructure 
  themes. These features could be:
  * school district information,
  * characteristics of the housing and rental stock as well as occupancy,
  * available employment opportunities etc.
* Add relevant geographic characteristics into the analysis such as the distances
  to the CBD or employment centers, road network or recreational amenities.
* Depending on data availability increase the spatial resolution and move from
  census tracts to block groups.
* Increase the number of clusters in order to boost the homogeneity of the market
  areas.
* Edit the spatial contiguity matrix to introduce physical boundaries between 
  market areas, for instance, tracts located on either side of Trinity river or 
  major highways.
  
# General modelling strategy

1. Select those _features_ which characterized the diversity among the spatial objects and
that are aligned with the objectives of the cluster analysis. The distribution of a selected variable should
be outstandingly different at least for one cluster from the other clusters. See the section "Inspect feature profiles for each cluster". Are two variable profiles similar across all clusters then these variables 
are redundant with respect to the cluster analysis and one variable should be removed
from the analysis.
2. Select that _spatial dissimilarity metric_ which leads to an interpretable regionalization 
without undermining the cluster homogeneity. Some spatial fragmentation of clusters is permitted.
3. First, identify the _number of clusters_ $K$ while ignoring the regionalization, i.e., use $\alpha=0$, by studying the dendogram and systematically increasing the number of clusters. Adopt that number of clusters which leads to interpretable results.
4. Second, add _regionalization_ to the optimal number of clusters by systematically increasing the mixture parameter $\alpha=0$. Select that solution, which reveals an interpretable regionalization and feature profiles of clusters. If necessary, some regions can be spatially fragmented. Remember, clusters should be as distinct as possible from each other with respect to the selected features but has homogeneous as possible within each cluster. A spatially fragmented regionalization simply means that there are more than one region which share a similar feature profile. 

___Homogeneity within the clusters and their unique profiles are more important than a spatially unfragmented regionalization.___

# Setting the data up

## Definition of the spatial dissimilarity metric

Set the spatial tessellation of the census tracts up and prepare the contiguity 
space dissimilarity among the census tracts.

```{r Adjacency, echo=TRUE, fig.height=7.2}
data(tractShp)
tractShp <- tractShp[tractShp$NIGHTPOP!=0, ]         # Remove 2 airport tracts with NA's

##
## Identify spatial neighboring census tracts and process them
##
nb <- spdep::poly2nb(tractShp, queen=F)              # extract first order neighbors links
B <- spdep::nb2mat(nb, style="B")                    # convert neighbor list to binary matrix

## Visualize first order neighbors
plot(tractShp, col="palegreen3", border=grey(0.9), axes=T) 
plot(nb, coords=coordinates(tractShp), pch=19, cex=0.1, col="blue", add=T)
title("Spatial Neighbors Links among Tracts") 

## Transform the contiguity matrix into spatial dissimilarity matrix
geoDist <- 1-B; 
diag(geoDist) <- 0; 
geoDist <- as.dist(geoDist)

##
## Alternative spatial dissimilarity matrices
##
# # Calculate shortest path distances between the tracts
# BNa <- ifelse(B==0,NA,B)                             # recode 0's to NA's
# allPath <- e1071::allShortestPaths(BNa)              # calucate the shortest path among all nodes
# topoDist <- as.dist(allPath$length)                  # number of steps from origin to destination node
# e1071::extractPath(allPath, 1, 527)                  # example: path between tract 1 and tract 527
#
# # spherical distances matrix among tracts in km
# sphDist <- as.dist(sp::spDists(tractShp, longlat=T)) 
```

## Definition of the feature dissimilarity metric

Set the variables of the census tracts up, impute 25 missing median home values 
and define the feature space dissimilarity matrix.

```{r DataPrep, echo=TRUE, fig.height=7.2, fig.width=7.2}
##
## Select variables for feature dissimilarity organized into three groups:
##   [a] demographic features
##   [b] socio-economic features
##   [c] housing infrastructure features
##

## Calculate additional features
tractShp$PRE1960 <- tractShp$PCTB1950+tractShp$PCTB1940+tractShp$PCTBPRE
tractShp$POST2000 <- tractShp$PCTB2000+tractShp$PCTB2010

## Feature list
varKeep <- c("PCTWHITE","PCTBLACK","PCTHISPAN","MEDAGE",
             "PCTUNIVDEG","HHMEDINC","PCTUNEMP","PCTNOHINS",
             "POPDEN","PCTDAYPOP","PCTHUVAC","MEDVALHOME","PRE1960","POST2000")
xVars <- tractShp@data
xVars <- xVars[varKeep]
summary(xVars)

## Replace missing values in MEDVALHOME by neighbors average
xVars <- VIM::kNN(xVars, variable="MEDVALHOME", k=5, imp_var=FALSE, numFun=median)
summary(xVars$MEDVALHOME)

## ID each row of xVars
row.names(xVars) <- 1:nrow(xVars)

## Calculate feature distance matrix
featDist <- dist(scale(xVars))
```

# Executing the cluster analysis

## Identification of the combined dissimilarity metric

Identify the $\alpha$ parameter which balances the convex mixture of the feature
space dissimilarities (black curve) with the contiguity space dissimilarities 
(red curve):

$$(1-\alpha)*FeatureDissimilarity + \alpha*ContiguityDissimilarity$$.

Preference should be given to the feature dissimilarity (black) and for neither 
dissimilarity metric the intra-cluster homogeneity should not drop rapidly for 
the sequence of $\alpha$-values. Low $\alpha$ values may lead to spatially split 
clusters.

For this prototype analysis the number of clusters is set to $K=12$, because $12$ 
is the maximum number of factor levels that the function "mapColorQual" can display. 
A work-around to increase the number of clusters is to split $K>12$ clusters 
across several maps of maximal $12$ factor levels. Please see step 8 below.

The selected mixture parameter $\alpha$ is set at $\alpha=0.2$ because for smaller
$\alpha$-values the intra-cluster homogeneity for spatial contiguity metric drops
rapidly.

```{r AlphaMixture, echo=TRUE, fig.height=7.2}
##
## Evaluate mixture of feature and spatial dissimilarity.
##
K <- 9                             # Number of distinct clusters
range.alpha <- seq(0, 1, by=0.1)   # Evaluation range
cr <- choicealpha(featDist, geoDist, range.alpha, K, graph=TRUE)
```

## Spatially constrained cluster analysis

Show the cluster generation history in a dendrogram and highlight the 12 market 
areas of census tracts.

```{r ClusterAnalysis, echo=TRUE, fig.height=7.2, fig.width=7.2}
## 
## Perform spatially constrained cluster analysis
##
tree <- hclustgeo(featDist, geoDist, alpha=0.2)
plot(tree, hang=-1)
rect.hclust(tree, k=K)
```

Evaluate the number of census tracts per market area.

```{r TractCount, echo=TRUE, fig.height=7.2}
## Number of census tracts per market area
neighClus <- as.factor(cutree(tree, K))        # Determine cluster membership
table(neighClus)                               # number of tracts in each cluster
```

# Evaluation of the identified market areas

Map the identified homogeneous market area. An interpretation can be
performed by their locations within Dallas County and by the levels and dispersion 
of their underlying features. 

Note that some internally homogeneous clusters of market areas are distributed 
disjunctly across Dallas County. This implies that areas of similar characteristics can be
found in several locations of Dallas County.

## Map identified market areas
```{r AllMap, echo=TRUE, fig.height=7.2}
##
## Map Results
##
mapColorQual(neighClus, tractShp, 
             map.title ="Spatially constrained Cluster Analysis",
             legend.title="Cluster\nId.", legend.cex=0.9)
plot(lakesShp, col="skyblue", border="skyblue",add=T)
plot(hwyShp, col="cornsilk2", lwd=4, add=T)
```

## Attribute analysis of the identified market areas

Evaluate the unique feature compositions for each market area. A market area 
should significantly differ from the remaining market areas in its variable 
profile by at least one feature or a combination of features.

To maintain a sufficient visual resolution, the variable box-plots by market areas
are split into two panels.

```{r AreaCharacteristics, echo=TRUE, fig.height=7.2, fig.width=7.2}
##
## Evaluate cluster characteristics
##
plotBoxesByFactor(xVars[,1:7], neighClus, ncol=2, zTrans=T, varwidth=F)
plotBoxesByFactor(xVars[,8:14], neighClus, ncol=2, zTrans=T, varwidth=F)
```

## Show heatmap of clusters by features

The heatmap shows the dendrogram of the clustered census tracts on the left margin and their associated cluster membership numbers on the right margin. In contrast, the features are grouped together based on their similarity and their associated dendrogram is shown on the top margin. In the heatmap the red color ramp denotes feature values for individual census tracts that are above average, whereas the blue color ramp denotes census tracts below average.

```{r Heatmap, fig.width=9, fig.height=10}
col <- rev(colorRampPalette(brewer.pal(8, "RdBu"))(8))

## Calculate hclust dendrograms
hclust_rows <- as.dendrogram(tree)  
hclust_cols <- as.dendrogram(hclust(dist(t(scale(xVars)))))

## Draw heatmap with hclust dendrograms
stats::heatmap(as.matrix(scale(xVars)), col=col,                                
               Rowv = hclust_rows, Colv = hclust_cols,
               labRow=as.character(cutree(tree, K)))
```

# Mapping a subset of clusters
To map just a set of selected clusters or, alternatively, for more than 12 
generated clusters a sequence of maps with consecutive subsets of clusters the 
remaining clusters are set to NA. Tract with NA's are displayed in light grey. 

Clusters 4 and 10 are predominant employment centers with a proportionally higher
number of day-time population relative to the nighttime population.

```{r SelectMap, echo=TRUE, fig.height=7.2}
## Select subset nLevels[c(4,8)] and set remaining clusters to NA
nLevels <- levels(neighClus)
neighClusSet <- as.factor(ifelse(neighClus %in% nLevels[c(4,8)], neighClus, NA))
mapColorQual(neighClusSet, tractShp, 
             map.title ="Clusters of Major Employment Centers",
             legend.title="Cluster\nId.", legend.cex=0.9)
plot(lakesShp, col="skyblue", border="skyblue",add=T)
plot(hwyShp, col="cornsilk2", lwd=4, add=T)
```
