---
title: "Spatial Interaction Models of Italian Provincial Migration"
author: "Michael Tiefelsdorf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Italy migration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r settings, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 5
)
```

## Introduction

This vignette demonstrates spatial interaction (gravity) modelling using internal
migration flows among the 95 provinces of Italy (1994). Three datasets from the
`TexMix` package are used:

- `ItalyShp`: spatial polygons and socioeconomic attributes for Italian provinces
- `Italy_neighborsShp`: background polygons for neighboring countries and territories
- `Italy_migration`: a 95 × 95 origin-destination matrix of inter-provincial
  migrant counts

The analysis proceeds in four stages:

1. **Exploratory mapping** of socioeconomic and migration patterns
2. **Flow mapping** of the dominant migration corridors
3. **Gravity models**: classical, over-dispersed, and doubly-constrained
4. **Augmented models** that add socioeconomic push-pull variables

```{r setup, message = FALSE, warning = FALSE}
library(TexMix)
library(sp)
library(spdep)
```

---

## Data

We load the three datasets and compute two spatial summaries used throughout:
the bounding box of Italy (for consistent map extents) and the matrix of
great-circle distances between provincial centroids.

```{r data}
data(ItalyShp)
data(Italy_neighborsShp)
data(Italy_migration)

bBox     <- sp::bbox(ItalyShp)
centroid <- sp::coordinates(ItalyShp)
dij      <- sp::spDists(centroid, longlat = TRUE)
diag(dij) <- NA   # self-distances are undefined
```

---

## Exploratory Mapping

### Regional structure

Italy's 95 provinces are grouped into three macro-regions (North, Central, and
South) which serve as fixed effects in the doubly-constrained model below.

```{r map-regions}
sp::plot(Italy_neighborsShp, axes = TRUE, col = grey(0.9), border = "white",
         xlim = bBox[1, ], ylim = bBox[2, ])
mapColorQual(as.factor(ItalyShp$REGION), ItalyShp,
             map.title    = "Regions of Italy",
             legend.title = "Region",
             add.to.map   = TRUE)
```

### Fertility rate

Total fertility rate varies substantially across the peninsula, with notably
lower rates in the North and Central regions.

```{r map-fertility}
sp::plot(Italy_neighborsShp, axes = TRUE, col = grey(0.9), border = "white",
         xlim = bBox[1, ], ylim = bBox[2, ])
#mapColorRamp(ItalyShp$TOTFERTRAT, ItalyShp, breaks = 7,
#             map.title    = "Italy Fertility Rate",
#             legend.title = "Fertility Rate",
#             add.to.map   = TRUE, legend.cex = 0.7)
```

### Net migration ratio

The log of the inflow-to-outflow ratio identifies provinces as net migration
destinations (positive values) or net sources (negative values).

```{r map-migration-ratio}
ItalyShp$logMigRatio <- log(ItalyShp$INFLOW / ItalyShp$OUTFLOW)

hist(ItalyShp$logMigRatio,
     main = "Log Migration Ratio",
     xlab = "log(Inflow / Outflow)")

cat("Net losers (ratio < 0):", sum(ItalyShp$logMigRatio <  0), "\n")
cat("Net gainers (ratio > 0):", sum(ItalyShp$logMigRatio >  0), "\n")
```

```{r map-bipolar}
sp::plot(Italy_neighborsShp, axes = TRUE, col = grey(0.9), border = "white",
         xlim = bBox[1, ], ylim = bBox[2, ])
mapBiPolar(ItalyShp$logMigRatio, ItalyShp,
           neg.breaks   = 3, pos.breaks = 5, break.value = 0,
           map.title    = "Italy Migration Ratio",
           legend.title = "Loss vs. Gain",
           add.to.map   = TRUE, legend.cex = 0.9)
```

---

## Migration Flow Matrix

### Flow matrix setup

We extract the 95 × 95 origin-destination matrix from `Italy_migration` and
set the diagonal to `NA` (a province does not migrate to itself).

```{r flow-matrix}
mij       <- as.matrix(Italy_migration[, 2:96])
diag(mij) <- NA
```

### Mapping dominant flows

The top 1% of flows reveal the dominant migration corridors between provinces.

```{r map-flows}
bigmig        <- mij > quantile(as.vector(mij), prob = 0.99, na.rm = TRUE)
diag(bigmig)  <- FALSE
prov.link     <- spdep::mat2listw(bigmig, style = "B", zero.policy = TRUE)$neighbours

sp::plot(Italy_neighborsShp, axes = TRUE, col = grey(0.9), border = "white",
         xlim = bBox[1, ], ylim = bBox[2, ])
sp::plot(ItalyShp, col = "palegreen3", border = grey(0.9), add = TRUE)
plot(prov.link, coords = centroid, pch = 19, cex = 0.1,
     arrows = TRUE, length = 0.05, col = "blue", add = TRUE)
title("Top 1% of Migration Flows")
box()
```

---

## Building the Origin-Destination Data Frame

`prepIJDf()` reshapes province-level attributes into a long-format OD data
frame with one row per origin-destination pair (95² = 9,025 rows). With
`logTrans = TRUE` and `zTrans = TRUE`, each positive numeric variable is
expanded into log-standardized origin (`.iLZ`), destination (`.jLZ`), and
log-ratio (`.ijLZ`) columns.

```{r prep-df}
dfShort <- ItalyShp@data[, c("PROVNAME", "REGION", "TOTPOP94", "TOTFERTRAT",
                              "FEMMARAGE9", "DIVORCERAT", "ILLITERRAT", "TELEPERFAM")]
dfShort$PROVNAME <- as.factor(dfShort$PROVNAME)
df <- prepIJDf(dfShort, mij = mij, dij = dij, logTrans = TRUE, zTrans = TRUE)
dim(df)
```

---

## Gravity Models

All models use a Poisson family and exclude self-pairs with
`subset = (ID.i != ID.j)`.

### Model 1: Classical gravity model

The baseline gravity model regresses flow counts on log-distance and the
log-standardized populations of the origin and destination provinces.

```{r mig1}
mig1.glm <- glm(mij ~ log(dij) + TOTPOP94.iLZ + TOTPOP94.jLZ,
                data   = df,
                family = poisson,
                subset = (ID.i != ID.j))
summary(mig1.glm)
```

**Over-dispersion check:** a Pearson dispersion ratio well above 1 indicates
that the Poisson variance assumption is too restrictive.

```{r mig1-dispersion}
sum(residuals(mig1.glm, type = "pearson")^2) / mig1.glm$df.residual
```

### Model 2: Quasi-Poisson

Quasi-Poisson keeps the same mean function but estimates a free dispersion
parameter rather than fixing variance equal to the mean. An important property
of Poisson-family models is that total predicted flow equals total observed flow.

```{r mig2}
mig2.glm <- glm(mij ~ log(dij) + TOTPOP94.iLZ + TOTPOP94.jLZ,
                data   = df,
                family = quasipoisson,
                subset = (ID.i != ID.j))
summary(mig2.glm)

cat("Observed total flow: ", sum(df$mij[df$ID.i != df$ID.j], na.rm = TRUE), "\n")
cat("Predicted total flow:", round(sum(predict(mig2.glm, type = "response"))),  "\n")
```

```{r mig2-fitplot}
plot(df$mij[df$ID.i != df$ID.j],
     predict(mig2.glm, type = "response"),
     log  = "xy",
     xlab = "Observed Migration Flows",
     ylab = "Predicted Migration Flows",
     main = "Quasi-Poisson: Observed vs. Predicted")
abline(a = 0, b = 1, col = "red")
```

### Model 3: Negative binomial

Negative binomial explicitly models the extra variance with a free parameter θ.
Unlike Poisson-family models, it does not guarantee that total predicted and
observed flows match.

```{r mig3}
mig3.glm <- MASS::glm.nb(mij ~ log(dij) + TOTPOP94.iLZ + TOTPOP94.jLZ,
                          data    = df,
                          subset  = (ID.i != ID.j),
                          control = glm.control(epsilon = 1e-8, maxit = 100))
summary(mig3.glm)

cat("Observed total flow: ", sum(df$mij[df$ID.i != df$ID.j], na.rm = TRUE), "\n")
cat("Predicted total flow:", round(sum(predict(mig3.glm, type = "response"))),  "\n")
```

### Model 4: Doubly-constrained model

A doubly-constrained model adds province-level fixed effects for both origins
and destinations, ensuring that predicted inflows and outflows match the
observed values at every province. Sum (deviation) contrasts center these
effects so that coefficients remain interpretable.

```{r mig4}
mig4.glm <- glm(mij ~ log(dij) + PROVNAME.iF + PROVNAME.jF,
                data      = df,
                family    = poisson,
                subset    = (ID.i != ID.j),
                contrasts = list(PROVNAME.iF = "contr.sum",
                                 PROVNAME.jF = "contr.sum"))
```

We verify the constraint: observed and predicted inflows should be identical at
every province.

```{r mig4-balance}
ObsInflow  <- tapply(df$mij,
                     df$PROVNAME.iF,
                     sum)
PredInflow <- tapply(predict(mig4.glm, type = "response"),
                     df$PROVNAME.iF[df$ID.i != df$ID.j],
                     sum)
head(cbind(Observed = ObsInflow, Predicted = round(PredInflow)), 10)
```

---

## Optimizing the Distance Decay Function

The log transformation of distance (λ = 0) may not be optimal. We profile the
log-likelihood over a grid of Box–Cox power parameters (λ ∈ [−1.4, −0.4]) to
find the best-fitting transformation.

```{r lambda-search}
# Replace NA self-distances with a small constant so power transforms are defined
df$dij <- ifelse(is.na(df$dij), 0.1, df$dij)

lambda <- seq(-1.4, -0.4, 0.01)
LR     <- numeric(length(lambda))

for (i in seq_along(lambda)) {
  fit    <- glm(mij ~ car::bcPower(dij, lambda[i]) + TOTPOP94.iLZ + TOTPOP94.jLZ,
                data = df, family = poisson, subset = (ID.i != ID.j))
  LR[i] <- as.numeric(logLik(fit))
}

lambdaMax <- lambda[which.max(LR)]
cat("Optimal lambda:", lambdaMax, "\n")
```

```{r lambda-plot}
plot(lambda, LR, type = "l",
     ylab = "Log-likelihood",
     xlab = expression(lambda),
     main = "Profile Log-Likelihood for Distance Transformation")
abline(v = lambdaMax, col = "red", lwd = 2, lty = 2)
text(lambdaMax + 0.02, min(LR) + diff(range(LR)) * 0.05,
     bquote(hat(lambda) == .(lambdaMax)),
     adj = 0, col = "red")
```

The optimal λ falls between −1 (inverse distance) and 0 (log distance),
indicating that neither standard transformation provides the best fit.

### Effect on distance distributions

```{r dist-boxplot, fig.height = 6}
iOrdProv <- reorder(df$PROVNAME.iF, df$dij, median)

boxplot(dij ~ iOrdProv, data = df, subset = (ID.i != ID.j),
        main     = "Inter-provincial Distances by Origin\n(before Box-Cox)",
        xlab     = "Province",
        ylab     = "Distance (km)",
        las      = 2,
        cex.axis = 0.45)

boxplot(car::bcPower(dij, lambdaMax) ~ iOrdProv, data = df, subset = (ID.i != ID.j),
        main     = "Inter-provincial Distances by Origin\n(after Box-Cox)",
        xlab     = "Province",
        ylab     = "Transformed Distance",
        las      = 2,
        cex.axis = 0.45)
```

---

## Augmented Models

We now add socioeconomic push-pull variables alongside the optimized distance
term. Two parameterizations are compared.

### Model 5: Separate origin and destination effects

Each variable enters as independent origin (`.iLZ`) and destination (`.jLZ`)
terms, allowing asymmetric push and pull effects.

```{r mig5}
mig5.glm <- glm(mij ~ car::bcPower(dij, lambdaMax) +
                  TOTPOP94.iLZ   + TOTPOP94.jLZ   +
                  TOTFERTRAT.iLZ + TOTFERTRAT.jLZ +
                  FEMMARAGE9.iLZ + FEMMARAGE9.jLZ +
                  DIVORCERAT.iLZ + DIVORCERAT.jLZ +
                  ILLITERRAT.iLZ + ILLITERRAT.jLZ +
                  TELEPERFAM.iLZ + TELEPERFAM.jLZ,
                data   = df,
                family = poisson,
                subset = (ID.i != ID.j))
summary(mig5.glm)
car::vif(mig5.glm)
```

```{r mig5-fitplot}
plot(df$mij[df$ID.i != df$ID.j],
     predict(mig5.glm, type = "response"),
     log  = "xy",
     xlab = "Observed Migration Flows",
     ylab = "Predicted Migration Flows",
     main = "Model 5: Augmented (Separate Origin / Destination)")
abline(a = 0, b = 1, col = "red")
```

### Model 6: Origin-to-destination ratio effects

This model replaces the separate origin and destination terms with the
log-ratio of origin to destination values (`.ijLZ`), which halves the number
of socioeconomic parameters and tests whether the direction of the differential
(rather than its absolute level) drives migration.

```{r mig6}
mig6.glm <- glm(mij ~ car::bcPower(dij, lambdaMax) +
                  TOTPOP94.iLZ    + TOTPOP94.jLZ    +
                  TOTFERTRAT.ijLZ + FEMMARAGE9.ijLZ +
                  DIVORCERAT.ijLZ + ILLITERRAT.ijLZ +
                  TELEPERFAM.ijLZ,
                data   = df,
                family = poisson,
                subset = (ID.i != ID.j))
summary(mig6.glm)
car::vif(mig6.glm)
```

```{r mig6-fitplot}
plot(df$mij[df$ID.i != df$ID.j],
     predict(mig6.glm, type = "response"),
     log  = "xy",
     xlab = "Observed Migration Flows",
     ylab = "Predicted Migration Flows",
     main = "Model 6: Augmented (Origin/Destination Ratio)")
abline(a = 0, b = 1, col = "red")
```
